Annotations on Quantum Pseudorandomness and $k$-Designs

### 2020-11-22 PHYS790/final.tex
Finished most of MC portion.

\usepackage{appendix}

@@

\begin{abstract}
  We investigate constructions of efficient approximate unitary $k$-designs or distributions over the unitary group which model the Haar distribution up to the $k^{th}$-moment. In particular, we review the paper by Harrow and Low as a directed reading conducted during the Spring 2020 semester.
\end{abstract}

@@

\section{Approximate 2-designs}

@@

      \item $\widehat{G}(p,a) = \frac{1}{d^2-1}, \quad a,p \neq 0$
      \begin{proof}
        As above, this results from $\widehat{G}$ acting on $\mathcal{F}$ along with the condition that it must send all other operators to zero.
      \end{proof}
    \end{enumerate}
    We conclude from the definition of $T(p,q)$ for $p,q \in \{0,\cdots,d^2-1\}$ that
    %
    \begin{equation}
      T(p,q) = \begin{cases}
        0 \quad & p \neq q \\
        \sigma_0 \otimes \sigma_0 & p = q = 0 \\
        \frac{1}{d^2 - 1}\sum_{\ell \neq 0}  \sigma_\ell \otimes \sigma_\ell & p = q \neq 0
      \end{cases}
    \end{equation}
\end{enumerate}
%
Recall that $T(p,q)$ refers to the expected \emph{state} from applying unitary $U \sim \mathbb{U}(d)$ to the basis state $\sigma_p \otimes \sigma_q$. From the determination above, it is simple to see that $U$ uniformly samples from all diagonal basis states when acting on $\sigma_p \otimes \sigma_p, \; p \neq 0$. Strikingly enough, applying $U \sim \mathbb{U}(d)$ conserves the total sum of the coefficients of the diagonal basis states.
%
\begin{lemma}
  The total sum of the coefficients $\sum_{p} \gamma_0(p,p)$ is conserved by $U \sim \mathbb{U}(d)$ for state $\rho = \frac{1}{d}\sum_{p_1,p_2} \gamma_0(p_1,p_2) \sigma_{p_1} \otimes \sigma_{p_2}$.
\end{lemma}
%
Normalizing this sum allows us to view it as a \emph{probability distribution}. Furthermore, the evolution of our random circuit scheme is amenable to Markov chain analysis. As an illustrative example, suppose we have an $n$-qubit system and so far have applied $t$ steps of our scheme. As per usual, we pick our $2$-qubit unitary $U \sim \mathbb{U}(4)$. Now if we were to pick qubits $1,2$, our random $2$-qubit unitary would only act on the first two qubits non-trivially whereas every other qubit would just transform via the identity map. The expected \emph{coefficients} for step $t+1$ will work out to be
%
\begin{align}
    \gamma_{t+1}(p_1,\cdots,p_n,q_1,\cdots,q_n) = \begin{cases}
       0 & (p_1 ,p_2) \neq (q_1, q_2) \\
       \gamma_t(0,0,\cdots, p_n, 0,0,\cdots,q_n ) & (p_1,p_2) = (q_1, q_2) = (0,0) \\
       \frac{1}{15}\sum_{\substack{r_1,r_2 \\ r_1r_2 \neq 00}} \gamma_t(r_1,r_2,\cdots p_n,r_1,r_2,\cdots,q_n) & (p_1,p_2) = (q_1,q_2) \neq (0,0)
    \end{cases}
\end{align}
 Oliveira et.al \cite{oliveira} had the insight to map this evolution to a Markov chain as follows: \newline
%
\begin{theorem} \label{haarevol}
  Let $M$ be a Markov chain governing the evolution of states under $\widehat{G}$ where $\widehat{G}$ samples from the Haar distribution over $\mathbb{U}(2^n)$. Then $M$ exhibits the following transitions:
  %
  \begin{enumerate}
    \item The states of the Markov chain are $\Omega = \{0,1,2,3\}^n$
    \item The transitions adhere to the guidelines below: Suppose we start with a state indexed by string $\alpha = \alpha_1\cdots\alpha_n$
      \begin{enumerate}
        \item Pick two distinct sites $i < j$ uniformly at random.
        \item If $i = j = 0$, remain at the same state. Otherwise pick $\beta_i\beta_j = \{0,1,2,3\}^2/\{00\}$ uniformly at random. Transition to the string
        $$\alpha_1\alpha_2,\cdots\alpha_{i-1}\beta_i\alpha_{i+1}\cdots \alpha_{j-1}\beta_{j}\alpha_{j+1}\cdots\alpha_n$$
      \end{enumerate}
  \end{enumerate}
\end{theorem}

\subsection{Convergence of the Random Circuit}
Before embarking on the Markov chain analysis, we must prove that our random circuit indeed converges to Haar distribution. This will be proven for the general case of a universal distribution (Definition (\ref{univdef})) on $\mathbb{U}(d)$. Specifcally, we will prove that any universal distribution on $\mathbb{U}(d)$ will be $k$-copy gapped (Definition (\ref{gappeddef})). In turn, this leads to convergence in sense of $\lim_{t \rightarrow \infty} (\widehat{G}_{\mu})^{t} = \widehat{G}_{\mathbb{U}(d)}$. To see this, observe that another formulation of a $k$-copy gapped distributon over $\mathbb{U}(d)$ is
\begin{equation} \label{gappeddis}
  ||G_{\mu} - G_{\mathbb{U}(d)}||_\infty < 1
\end{equation}
where $G_{\mu} = \mathbb{E}_{U \sim \mu} \left[ U^{\otimes k} \otimes (U^\dagger)^{\otimes k} \right]$. To justify this, first recall that $\widehat{G}$ and $G$ are unitarily-related by Definition \ref{Tdef}, so in proving \ref{gappeddis}, we can argue on the operators $\widehat{G}$ instead. In view of this, $\widehat{G}_{\mathbb{U}(d)}$ is a projector and thus has eigenvalues 0 and 1. If we can prove that any eigenvector of $\widehat{G}_{\mathbb{U}(d)}$ of eigenvalue one is also an eigenvector of $\widehat{G}_{\mu}$ of eigenvalue one i.e the $k$-symmetric subspace of $(\mathbb{C}^d)^{\otimes k}$ is also fixed by $\widehat{G}_{\mu}$. Thus, Equation $\ref{gappeddis}$ shows that indeed we can converge to $G_{\mathbb{U}(d)}$ by iterating our scheme (as denoted by $(\widehat{G}_\mu)^t$). Note that $\widehat{G}_{\mu}$ may not even be Hermitian, so we cannot directly argue on its spectrum (as the spectrum may not completely contain real values).
%
\begin{theorem} \label{convergence}
  Let $\mu$ be a universal distribution over $\mathbb{U}(d)$. Then all eigenvectors of $G_{\mathbb{U}(d)}$ of eigenvalue one will also be eigenvectors of $G_{\mu}$ of eigenvalue one. Furthermore, $\mu$ will be $k$-copy gapped.
\end{theorem}
%
\noindent For the sake of presentation, the proof of this will be delegated towards the appendix of this report.
%
\subsection{Markov Chain Analysis}
As discussed in a previous section, the central idea of proving \emph{rapid} convergence involves mapping the evolution of the inital state under our random circuit to that of a Markov chain's. In fact, we can explicitly determine the operator $P$ capturing our dynamics. Let $\widehat{G}^{(ij)}$ denote the unitary transformation resulting from sampling some unitary $U \sim \mathbb{U}(d)$ and applying $U$ to qubits $i$ and $j$ while leaving all other qubits undisturbed.
%
\begin{equation} \label{schemechain}
P = \frac{1}{n(n-1)}\sum_{i \neq j} \widehat{G}^{(ij)}
\end{equation}
as evidenced by the equality
\begin{equation}
  \gamma_{t+1}(\textbf{p}) = \frac{1}{n(n-1)}\sum_{i \neq j}\sum_{\textbf{q}} \widehat{G}^{(ij)}(\textbf{p},\textbf{q})\gamma_t(\textbf{q})
\end{equation}
where $\textbf{p},\textbf{q} \in \{0,1,2,3\}^2$ and $\gamma_t(\textbf{p})$ represents the expected coeffcients at timestep $t$.
%
A major caveat of interpreting $P$ as a \emph{transition matrix} for some suitable Markov chain on the diagonal states $\sigma_p \otimes \sigma_p$ is the possibility that our initial coefficients of diagonal states $\gamma_0(p,p)$ may not even sum to one or even be non-negative. It turns out that this doesn't bring too much trouble into the analysis as pointed out by Corollary \ref{maincorollary}. For now, let us tackle the case where the two conditions above are true and direct our focus on the evolution of the diagonal states. Of course, we need to also show that the expected coeffcients for the non-diagonal states $\gamma_0(p,q)$ vanish rapidly as well. These matters will be addressed in the upcoming section.

\subsubsection{Preliminaries on Markov Chains}
Let us introduce some defintions of properties of the Markov chains we are interested in. Although we assume familiarity with the definition of a Markov chain along with the its associated transition matrix $P$, we refer the reader to the reference \cite{cover} for greater detail. Recall that if $q^{(0)}$ is some initial  distribution over the $\Omega$ then $q^{(t)} = q^{(0)}P^t$. The \emph{stationary distribution} of a Markov chain $\pi$ is a probability distribution over $\Omega$ such that $\pi P = \pi$.
%
\begin{definition}
  A Markov chain is said to be \emph{irreducible} if there exists a path $s \rightsquigarrow j$ for every pair of states $s,j \in \Omega$.
\end{definition}
%
\begin{definition}
  A Markov chain is said to be \emph{periodic} at state $i \in \Omega$ if there exists a positive integer $T > 1$ such for that some initial distribution $q^(0)$, if $q_i^{(t)} > 0$ then $t \in \{a + Tm \mid m \in \mathbb{N}\}$. A Markov chain is deemed as \emph{aperiodic} if none of its states are periodic. Equivaently, a Markov chain is aperiodic if all of its states contain self-loops.
\end{definition}
%
\begin{definition}
  A Markov chain is \emph{ergodic} if it is aperiodic and irreducible.
\end{definition}
\noindent As the below theorem reveals, ergodicity shows that any initial distribution over $\Omega$ converges to a unique stationary distribution.
%
\begin{theorem}
  If a Markov chain is ergodic, then it converges to a unique stationary distributon $\pi$ i.e $P$ has a unique eigenvector with eigenvalue one.
\end{theorem}
%
\noindent One can see this through a special case of the \emph{Perron-Frobenius Theorem for Regular Matrices}
%
\begin{theorem}
  Suppose that $P$ is transition matrix with positive entries and also that $P_{ij}^{k} > 0$ for some sufficiently large $k$ and for all entries $i,j$ (Regularity). Then the following must hold:
  \begin{enumerate}
    \item The largest eigenvalue must be real and postive
    \item The largest left eigenvector $\pi$ must be positive and unique
    \item $\lim_{k \rightarrow \infty} qP^{(k)} = \pi$ for all initial distributions $q$.
  \end{enumerate}
\end{theorem}
\noindent Since our matrices are stochastic i.e $\sum_{j} P_{ij} = 1$ for all $i$, our largest eigenvalue will be one. Furthermore, the left eigenvector of eigenvalue one will be the stationary distribution $\pi$. Since our Markov chain is both irreducible and aperiodic, it must be regular.
%Mention counterexamples if time remains
%
\begin{definition}
A Marko chain is \emph{reversible} if the following equality holds for all states $i,j \in \Omega$
\begin{equation}
  \pi(i)P_{ij} = \pi(j)P_{ji}
\end{equation}
\end{definition}
\noindent If the stationary distributon is uniform, then it is simple to see that reversability is equivalent to $P$ being symmetric. Reversability is necessary to invoke certain mixing time lemmas which are introduced in the upcoming section.

@@

Finally, one more significant tool is required to lower bound the spectral gap of our matrix $P$ in Equation \ref{schemechain}. It is worth stressing that the Markov chain associated to $P_{\mu}$ which we will define as
%
\begin{equation}
P_\mu - \frac{1}{n(n-1)}\sum_{i\neq j}\widehat{G_\mu}^{(ij)}
\end{equation}
For this chain, we sample from a $2$-copy gapped distribution $\mu $ over $\mathbb{U}(4)$ rather than over Haar distribution. It is unclear exactly how this chain evolves as we cannot explicitly compute $P_\mu$ as we did for $P_{\mathbb{U}(d)}$. However, with the \emph{Comparision Theorem}, we can at least lower bound the spectral gap associated with $P_\mu$, say $\Delta_\mu$, with $\Delta_{\mathbb{U}(d)}$. We first state a more general version of the theorem
%
\begin{theorem}[Refer to \cite{montenegro}, Theorem 2.3] \label{comptheorem}
  Let $P$ and $\widehat{P}$ be two Markov chains with the same state space $\Omega$ and the same stationary distribution $\pi$. Then, for ever $x \neq y \in \Omega$ with $\widehat{P}_{xy} > 0$, construct a set of directed paths $\mathcal{P}_{xy}$ from $x$ to $y$ along edges in $P$. Let $\mathcal{P} = \bigsqcup_{x\neq y} \mathcal{P}_{xy}$. We define the flow function $f: \mathcal{P} \rightarrow [0,1]$ such that
  \begin{equation}
    \sum_{\gamma_{xy} \in \mathcal{P}_{xy}} f(\gamma_{xy}) = \widehat{P}_{xy}
  \end{equation}
  then
  $$ \Delta \geq \widehat{\Delta}/A $$
  where
  $$ A = \max_{a \neq b, P_{ab} > 0}\frac{1}{\pi_aP_{ab}} \sum_{x \neq y}\sum_{
                          \substack {
                          \gamma_{xy} \in \mathcal{P}_{xy} \\
                          (a,b) \in \gamma_{xy}}}  \pi(x)f(\gamma_{xy})|\gamma_{xy}|$$
  where $|\gamma_{xy}|$ refers to the length of the path $\gamma_{xy}$.
\end{theorem}
\noindent Note that the bounds yielded will depend on the paths chosen for all $x \neq y \in \Omega$ such that $\widehat{P}_{xy} > 0$ and the values assigned to those paths by the constructed flow function $f$.
\newline

@@

\begin{theorem}
  Let $\mu$ be any 2-copy gapped distribution of gates. If $\mu$ is invariant under Hermitian conjugation then let $\Delta_\mu$ be the spectral gap of resulting Markov chain $P_\mu$. Then
  \begin{equation}
    \Delta_\mu = \Omega(\Delta_{\mathbb{U}(4)})
  \end{equation}
  If $\mu$ is not invariant under Hermitian conjugation, then let $\Delta_{PP^*}$ be the spectral gap of the chain $PP^*$. Then
  \begin{equation}
    \Delta_{PP^*} = \Omega(\Delta_{\mathbb{U}(4)})
  \end{equation}
\end{theorem}
\begin{proof}
  We only show the symmetric case for now. \newline

@@

A significant portion of \cite{harrlow} is dedicated to proving that $\Delta_{\mathbb{U}(4)} = \Theta(1/n)$. Invoking Lemma \ref{spectralgap} and \ref{2spectralgap} along with this tight bound on the spectral gap of the $\mathbb{U}(4)$ chain shows that
%
\begin{corollary}
  The chain $P_\mu$ has 1-norm mixing time $\mathcal{O}(n(n+\log1/\epsilon))$ and 2-norm $\mathcal{O}(n\log\frac{n}{\epsilon})$
\end{corollary}

@@

\newpage
\bibliographystyle{amsplain}
\bibliography{biblio}

>>>>

\usepackage{appendix}
\usepackage{tikz}
\usetikzlibrary{automata, positioning}

@@

\begin{abstract}
  We investigate constructions of efficient approximate unitary $k$-designs which are defined as probability distributions over the unitary group which model the Haar distribution up to the $k^{th}$-moment. In particular, we review the paper by Harrow and Low as a directed reading conducted during the Spring 2020 semester.
\end{abstract}

@@

\section{Approximate 2-designs via Local Random Quantum Circuits}

@@

      \item $\widehat{G}(p,a) = \frac{1}{d^2-1}, \quad a,p \neq 0$
      \begin{proof}
        As above, this results from $\widehat{G}$ acting on $\mathcal{F}$ along with the condition that it must send all other operators to zero.
    We conclude from the definition of $T(p,q)$ for $p,q \in \{0,\cdots,d^2-1\}$ that
    \begin{equation} \label{concreteT}
      T(p,q) = \begin{cases}
        0 \quad & p \neq q \\
        \sigma_0 \otimes \sigma_0 & p = q = 0 \\
        \frac{1}{d^2 - 1}\sum_{\ell \neq 0}  \sigma_\ell \otimes \sigma_\ell & p = q \neq 0
      \end{cases}
Recall that $T(p,q)$ refers to the expected \emph{state} from applying unitary $U \sim \mathbb{U}(d)$ to the basis state $\sigma_p \otimes \sigma_q$. From the determination above, it is simple to see that $U$ uniformly samples from all diagonal basis states when acting on $\sigma_p \otimes \sigma_p, \; p \neq 0$. Strikingly enough, applying $U \sim \mathbb{U}(d)$ conserves the total sum of the coefficients of the diagonal basis states.
\begin{lemma}
  The total sum of the coefficients $\sum_{p} \gamma_0(p,p)$ is conserved by $U \sim \mathbb{U}(d)$ for state $\rho = \frac{1}{d}\sum_{p_1,p_2} \gamma_0(p_1,p_2) \sigma_{p_1} \otimes \sigma_{p_2}$.
\end{lemma}
Normalizing this sum allows us to view it as a \emph{probability distribution}. Furthermore, the evolution of our random circuit scheme is amenable to Markov chain analysis. As an illustrative example, suppose we have an $n$-qubit system and so far have applied $t$ steps of our scheme. As per usual, we pick our $2$-qubit unitary $U \sim \mathbb{U}(4)$. Now if we were to pick qubits $1,2$, our random $2$-qubit unitary would only act on the first two qubits non-trivially whereas every other qubit would just transform via the identity map. By the explicit calculation shown in Equation \ref{concreteT}, the expected \emph{coefficients} for step $t+1$ will work out to be
\begin{align} \label{evolexample}
    \gamma_{t+1}(p_1,\cdots,p_n,q_1,\cdots,q_n) = \begin{cases}
       0 & (p_1 ,p_2) \neq (q_1, q_2) \\
       \gamma_t(0,0,\cdots, p_n, 0,0,\cdots,q_n ) & (p_1,p_2) = (q_1, q_2) = (0,0) \\
       \frac{1}{15}\sum_{\substack{r_1,r_2 \\ r_1r_2 \neq 00}} \gamma_t(r_1,r_2,\cdots p_n,r_1,r_2,\cdots,q_n) & (p_1,p_2) = (q_1,q_2) \neq (0,0)
    \end{cases}
 Oliveira et.al \cite{oliveira} had the insight to map the evolution of the $\gamma_t(p,p)$ coeffcients to a Markov chain as follows:
\begin{theorem} \label{haarevol}
  Let $M$ be a Markov chain governing the evolution of states under $\widehat{G}$ where $\widehat{G}$ samples from the Haar distribution over $\mathbb{U}(2^n)$. Then $M$ exhibits the following transitions:
    \item The states of the Markov chain are $\Omega = \{0,1,2,3\}^n$
    \item The transitions adhere to the guidelines below: Suppose we start with a state indexed by string $\alpha = \alpha_1\cdots\alpha_n$
        \item Pick two distinct sites $i < j$ uniformly at random.
        \item If $i = j = 0$, remain at the same state. Otherwise pick $\beta_i\beta_j = \{0,1,2,3\}^2/\{00\}$ uniformly at random. Transition to the string
        $$\alpha_1\alpha_2,\cdots\alpha_{i-1}\beta_i\alpha_{i+1}\cdots \alpha_{j-1}\beta_{j}\alpha_{j+1}\cdots\alpha_n$$
In the context of the example above, if we start with the coeffcient $\gamma_t(\textbf{q},\textbf{q})$ for $q \in \{0,1,2,3\}^n$, the only two possibilities are $(\textbf{q},\textbf{q}) = (0,0)$ and $(\textbf{q},\textbf{q}) \neq (0,0)$ since it is impossible for the coefficient to vanish by the first case in Equation \ref{evolexample}. Thus, the transitions presented in Theorem \ref{haarevol} indeed lead to the correct expected coefficient value $\gamma_{t+1}(\textbf{p},\textbf{p})$. Note that this is not the complete value for $\gamma_{t+1}$ as we still have to take the expectation over all the remaining qubit pairs.

\subsection{Convergence of the Random Circuit}
Before embarking on the Markov chain analysis, we must prove that our random circuit indeed converges to Haar distribution. This will be proven for the general case of a universal distribution (Definition (\ref{univdef})) on $\mathbb{U}(d)$. Specifcally, we will prove that any universal distribution on $\mathbb{U}(d)$ will be $k$-copy gapped (Definition (\ref{gappeddef})). In turn, this leads to convergence in sense of $\lim_{t \rightarrow \infty} (\widehat{G}_{\mu})^{t} = \widehat{G}_{\mathbb{U}(d)}$. To see this, observe that another formulation of a $k$-copy gapped distributon over $\mathbb{U}(d)$ is
\begin{equation} \label{gappeddis}
  ||G_{\mu} - G_{\mathbb{U}(d)}||_\infty < 1
\end{equation}
where $G_{\mu} = \mathbb{E}_{U \sim \mu} \left[ U^{\otimes k} \otimes (U^\dagger)^{\otimes k} \right]$. To justify this, first recall that $\widehat{G}$ and $G$ are unitarily-related by Definition \ref{Tdef}, so in proving \ref{gappeddis}, we can argue on the operators $\widehat{G}$ instead. In view of this, $\widehat{G}_{\mathbb{U}(d)}$ is a projector and thus has eigenvalues 0 and 1. If we can prove that any eigenvector of $\widehat{G}_{\mathbb{U}(d)}$ of eigenvalue one is also an eigenvector of $\widehat{G}_{\mu}$ of eigenvalue one i.e the $k$-symmetric subspace of $(\mathbb{C}^d)^{\otimes k}$ is also fixed by $\widehat{G}_{\mu}$. Thus, Equation $\ref{gappeddis}$ shows that indeed we can converge to $G_{\mathbb{U}(d)}$ by iterating our scheme (as denoted by $(\widehat{G}_\mu)^t$). Note that $\widehat{G}_{\mu}$ may not even be Hermitian, so we cannot directly argue on its spectrum (as the spectrum may not completely contain real values).
%
\begin{theorem} \label{convergence}
  Let $\mu$ be a universal distribution over $\mathbb{U}(d)$. Then all eigenvectors of $G_{\mathbb{U}(d)}$ of eigenvalue one will also be eigenvectors of $G_{\mu}$ of eigenvalue one. Furthermore, $\mu$ will be $k$-copy gapped.
\end{theorem}
%
\noindent For the sake of presentation, the proof of this will be delegated towards the appendix of this report.
%
\subsection{Markov Chain Analysis}
As discussed in a previous section, the central idea of proving \emph{rapid} convergence involves mapping the evolution of the inital state under our random circuit to that of a Markov chain's. In fact, we can explicitly determine the operator $P$ capturing our dynamics. Let $\widehat{G}^{(ij)}$ denote the unitary transformation resulting from sampling some unitary $U \sim \mathbb{U}(d)$ and applying $U$ to qubits $i$ and $j$ while leaving all other qubits undisturbed.
%
\begin{equation} \label{schemechain}
P = \frac{1}{n(n-1)}\sum_{i \neq j} \widehat{G}^{(ij)}
\end{equation}
as evidenced by the equality
\begin{equation}
  \gamma_{t+1}(\textbf{p}) = \frac{1}{n(n-1)}\sum_{i \neq j}\sum_{\textbf{q}} \widehat{G}^{(ij)}(\textbf{p},\textbf{q})\gamma_t(\textbf{q})
\end{equation}
where $\textbf{p},\textbf{q} \in \{0,1,2,3\}^2$ and $\gamma_t(\textbf{p})$ represents the expected coeffcients at timestep $t$.
%
A major caveat of interpreting $P$ as a \emph{transition matrix} for some suitable Markov chain on the diagonal states $\sigma_p \otimes \sigma_p$ is the possibility that our initial coefficients of diagonal states $\gamma_0(p,p)$ may not even sum to one or even be non-negative. It turns out that this doesn't bring too much trouble into the analysis as pointed out by Corollary \ref{maincorollary}. For now, let us tackle the case where the two conditions above are true and direct our focus on the evolution of the diagonal states. The ultimate goal of this line of thought will be to first prove Lemma \ref{mainlemma} through a carefully-done investigation of the \emph{mixing time} of $P$. Of course, we need to also show that the expected coeffcients of the non-diagonal states $\gamma_0(p,q)$ vanish rapidly as well. These matters will be addressed in a future section.

\subsubsection{Preliminaries on Markov Chains}
Let us introduce some defintions of properties of the Markov chains we are interested in. Although we assume familiarity with the definition of a Markov chain along with the its associated transition matrix $P$, we refer the reader to the reference \cite{cover} for greater detail. Recall that if $q^{(0)}$ is some initial  distribution over the $\Omega$ then $q^{(t)} = q^{(0)}P^t$. The \emph{stationary distribution} of a Markov chain $\pi$ is a probability distribution over $\Omega$ such that $\pi P = \pi$.
%
\begin{definition}
  A Markov chain is said to be \emph{irreducible} if there exists a path $s \rightsquigarrow j$ for every pair of states $s,j \in \Omega$.
\end{definition}
%
\begin{definition}
  A Markov chain is said to be \emph{periodic} at state $i \in \Omega$ if there exists a positive integer $T > 1$ such for that some initial distribution $q^(0)$, if $q_i^{(t)} > 0$ then $t \in \{a + Tm \mid m \in \mathbb{N}\}$. A Markov chain is deemed as \emph{aperiodic} if none of its states are periodic. Equivaently, a Markov chain is aperiodic if all of its states contain self-loops.
\end{definition}
%
\begin{definition}
  A Markov chain is \emph{ergodic} if it is aperiodic and irreducible.
\end{definition}
\noindent As the below theorem reveals, ergodicity shows that any initial distribution over $\Omega$ converges to a unique stationary distribution.
%
\begin{theorem}
  If a Markov chain is ergodic, then it converges to a unique stationary distributon $\pi$ i.e $P$ has a unique eigenvector with eigenvalue one.
\end{theorem}
%
\noindent One can see this through a special case of the \emph{Perron-Frobenius Theorem for Regular Matrices}
%
\begin{theorem} \label{perron}
  Suppose that $P$ is transition matrix with positive entries and also that $P_{ij}^{k} > 0$ for some sufficiently large $k$ and for all entries $i,j$ (Regularity). Then the following must hold:
  \begin{enumerate}
    \item The largest eigenvalue must be real and postive
    \item The largest left eigenvector $\pi$ must be positive and unique
    \item $\lim_{k \rightarrow \infty} qP^{(k)} = \pi$ for all initial distributions $q$.
  \end{enumerate}
\end{theorem}
\noindent Since our matrices are stochastic i.e $\sum_{j} P_{ij} = 1$ for all $i$, our largest eigenvalue will be one. Furthermore, the left eigenvector of eigenvalue one will be the stationary distribution $\pi$. Since our Markov chain is both irreducible and aperiodic, it must be regular as well.
%
\begin{remark}
  It is not difficult to find counterexamples to Theorem \ref{perron} if the Markov chain is not ergodic. For instance, consider the chain

  \begin{figure}[h!]
    \centering
    \begin{tikzpicture}
      \node[state]  (one) {1};
      \node[state, right=of one] (two) {2};

      \draw[every loop]
        (one) edge[loop above] node {1} (one)
        (two) edge[loop above] node {1} (two);
    \end{tikzpicture}
    \caption{An aperiodic but reducible chain}
  \end{figure}
  The transition matrix will just be $P = I$. Although the largest eigenvalue is positive and real, its associated left eigenvector is not unique. Thus, it does not have an unique stationary distribution. Another simple chain is

  \begin{figure}[h!]
    \centering
    \begin{tikzpicture}
      \node[state]  (one) {1};
      \node[state, right=of one] (two) {2};

      \draw[every loop]
        (one) edge[bend right, auto=right] node {1} (two)
        (two) edge[bend right, auto=right] node {1} (one);
    \end{tikzpicture}
    \caption{An irreducible but periodic chain}
  \end{figure}

  Once again, this chain does not have a unique left eigenvector of unit eigenvalue. Although the stationary distribution $\pi = [\frac{1}{2} \; \frac{1}{2}]$ is unique, no initial distribution converges to it besides $\pi$ itself.

  \begin{figure}[h!]
    \centering
    \begin{tikzpicture}
      \node[state]  (one) {1};
      \node[state, right=of one] (two) {2};

      \draw[every loop]
        (one) edge[loop above] node {1/2} (one)
        (one) edge[bend right, auto=right] node {1/2} (two)
        (two) edge[loop above] node {1} (two);
    \end{tikzpicture}
    \caption{An aperiodic, reducible chain which converges}
  \end{figure}
  The transition matrix will be $$ P = \left[ \begin{matrix} \frac{1}{2} & \frac{1}{2} \\
  0 & 1 \end{matrix} \right]$$ which has eigenvalues $\lambda_1 = 1, \lambda_2 = \frac{1}{2}$. Every initial distribution also converges to the unique stationary distribution $\pi = [0 \; 1]$. Hence, this chain exhibits a counterexample to a converse of Theorem \ref{perron}.
\end{remark}
%
%
%Mention counterexamples if time remains
%
\begin{definition}
A Markov chain is \emph{reversible} if the following equality holds for all states $i,j \in \Omega$
\begin{equation}
  \pi(i)P_{ij} = \pi(j)P_{ji}
\end{equation}
\end{definition}
\noindent If the stationary distributon is uniform, then it is simple to see that reversability is equivalent to $P$ being symmetric. Reversability is necessary to invoke certain mixing time lemmas which are introduced in the upcoming section.

@@

Finally, one more integral tool is required to lower bound the spectral gap of our matrix $P$ in Equation \ref{schemechain}. It is worth stressing that the Markov chain associated to $P_{\mu}$ which we will define as
%
\begin{equation}
P_\mu - \frac{1}{n(n-1)}\sum_{i\neq j}\widehat{G_\mu}^{(ij)}
\end{equation}
For this chain, we sample from a $2$-copy gapped distribution $\mu $ over $\mathbb{U}(4)$ rather than over Haar distribution. It is unclear exactly how this chain evolves as we cannot explicitly compute $P_\mu$ as we did for $P_{\mathbb{U}(d)}$. However, with the \emph{Comparision Theorem}, we can at least lower bound the spectral gap associated with $P_\mu$, say $\Delta_\mu$, with $\Delta_{\mathbb{U}(d)}$. We first state a more general version of the theorem
%
\begin{theorem}[Refer to \cite{montenegro}, Theorem 2.3] \label{comptheorem}
  Let $P$ and $\widehat{P}$ be two Markov chains with the same state space $\Omega$ and the same stationary distribution $\pi$. Then, for ever $x \neq y \in \Omega$ with $\widehat{P}_{xy} > 0$, construct a set of directed paths $\mathcal{P}_{xy}$ from $x$ to $y$ along edges in $P$. Let $\mathcal{P} = \bigsqcup_{x\neq y} \mathcal{P}_{xy}$. We define the flow function $f: \mathcal{P} \rightarrow [0,1]$ such that
  \begin{equation}
    \sum_{\gamma_{xy} \in \mathcal{P}_{xy}} f(\gamma_{xy}) = \widehat{P}_{xy}
  \end{equation}
  then
  $$ \Delta \geq \widehat{\Delta}/A $$
  where
  $$ A = \max_{a \neq b, P_{ab} > 0}\frac{1}{\pi_aP_{ab}} \sum_{x \neq y}\sum_{
                          \substack {
                          \gamma_{xy} \in \mathcal{P}_{xy} \\
                          (a,b) \in \gamma_{xy}}}  \pi(x)f(\gamma_{xy})|\gamma_{xy}|$$
  where $|\gamma_{xy}|$ refers to the length of the path $\gamma_{xy}$.
\end{theorem}
\noindent Note that the bounds yielded will depend on the paths chosen for all $x \neq y \in \Omega$ such that $\widehat{P}_{xy} > 0$ and the values assigned to those paths by the constructed flow function $f$.
\newline

@@

\begin{theorem} \label{gapbound}
  Let $\mu$ be any 2-copy gapped distribution of gates. If $\mu$ is invariant under Hermitian conjugation then let $\Delta_\mu$ be the spectral gap of resulting Markov chain $P_\mu$. Then
  \begin{equation}
    \Delta_\mu = \Omega(\Delta_{\mathbb{U}(4)})
  \end{equation}
  If $\mu$ is not invariant under Hermitian conjugation, then let $\Delta_{PP^*}$ be the spectral gap of the chain $PP^*$. Then
  \begin{equation}
    \Delta_{PP^*} = \Omega(\Delta_{\mathbb{U}(4)})
  \end{equation}
\end{theorem}
\begin{proof}
  We only show the symmetric case for now. \newline

@@

A significant portion of \cite{harrlow} is dedicated to proving that $\Delta_{\mathbb{U}(4)} = \Theta(1/n)$. Invoking Lemma \ref{spectralgap} and \ref{2spectralgap} along with this tight bound on the spectral gap of the $\mathbb{U}(4)$ chain shows that
%
\begin{corollary} \label{muchainmix}
  The chain $P_\mu$ has 1-norm mixing time $\mathcal{O}(n(n+\log1/\epsilon))$ and 2-norm $\mathcal{O}(n\log\frac{n}{\epsilon})$
\end{corollary}
In fact, by Theorem \ref{gapbound} and Lemma \ref{spectralgap}, the mixing time of the $P_{\mathbb{U}(4)}$ chain upper-bounds the mixing time of the $P_{\mu}$ chain. This mixing time bound is required to finally prove Lemma \ref{mainlemma} as intended.

@@

%Finally, we state the proof of Theorem \ref{maintheorem}. \newline
%\begin{proof}[(Proof of Theorem \ref{maintheorem})]
%\end{proof}

\bibliographystyle{amsplain}
\bibliography{biblio}
